---
title: "Spain Population Data Download"
author: "Angelo Santana"
date: "11/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

This file implements the code necessary for downloading the population data of Spain during the years 2008-2019 grouped by sex, age and autonomous community.

\  

# Data source

Population data of Spain can be downloaded from the [Spanish Statistical Office INE (_Instituto Nacional de Estadística_)](https://www.ine.es/en/) in the section [Demography and population > Population figures and Demographic Censuses > Population figures](https://www.ine.es/dyngs/INEbase/en/operacion.htm?c=Estadistica_C&cid=1254736176951&menu=ultiDatos&idp=1254735572981). As stated in this web page:

>> The __Population Figures__ provide a quantitative measurement of the population resident in Spain, in each Autonomous Community, in each province and on each island (in the island province), broken down according to basic demographic characteristics, such as sex, year of birth, age, nationality and country of birth. The population series is obtained from the intercensus population estimates for the 1971-2012 period, and from the Population Figures operation from 2012 onward.

>> This data is used as reference population figures in all of the statistical operations of the INE (surveys, National Accounts, indicators, etc.) and is transmitted on an international level as the official population data for Spain, for all intents and purposes.

\  

Data can be downloaded in several formats (excel, csv, PC-Axis, Json or plain text) in the section [Results](https://www.ine.es/dyngs/INEbase/en/operacion.htm?c=Estadistica_C&cid=1254736176951&menu=resultados&idp=1254735572981) of this page. We will use the R library `pxR` that allows us to directly download data from the web in pc-Axis format, which is an INE's own format. 

\  

## Population data downloading and tidying 

Direct access to data in Pc-Axis format is through the link [https://www.ine.es/jaxiT3/files/t/es/px/10262.px]:

```{r}
data <- pxR::read.px("https://www.ine.es/jaxiT3/files/t/es/px/10262.px")
```

The following code reads the data and prepares it in a tidy format for further processing with R:

```{r}
# Codes and names of Autonomous Communities in several languages
cacodes <- readxl::read_excel("CCAA codes.xlsx") %>% 
  select(codigoCA,ca=CA_INE)

# Resident population
population7120 <- data %>% 
  as.data.frame() %>% 
  filter(Comunidades.y.ciudades.autónomas!="Total Nacional", Edad!="Total",
         Sexo!="Ambos sexos") %>% 
  mutate(edad=gsub("[a-z ñ]","",Edad),
         edad=gsub("100á","100",edad),
         edad=gsub("85á","-1",edad),
         edad=as.numeric(edad)) %>% 
  filter(edad<85) %>% 
  rename(cca=Comunidades.y.ciudades.autónomas,
         sexo=Sexo,poblacion=value, periodo=Periodo) %>% 
  mutate(cca=as.character(cca),
         sexo=as.character(sexo),
         periodo=as.Date(periodo, format="%d de %b de %Y"),
         Edad=as.character(Edad),
         edad=ifelse(edad==-1,85,edad),
         grEdad=ifelse(edad<=50,"edad 0-49",
                       ifelse(edad<=64,"edad 50-64",
                              ifelse(edad<=74,"edad 65-74","edad 75 o más"))),
         año=lubridate::year(periodo),
         mesReferencia=lubridate::month(periodo),
         codigoCA=as.numeric(substr(cca,1,2))) %>% 
  full_join(cacodes) %>% 
  select(-c(cca)) %>% 
  group_by(año,periodo,mesReferencia,codigoCA,ca,sexo,grEdad) %>% 
  summarize(poblacion=sum(poblacion)) %>% 
  ungroup()

save(population7120,file="rdata/population7120.rdata")
```

